Derive final annotations based on NBAtlas reference

### 2025-08-19 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd

```{r}
final_annotation_df <- annotation_df |> 
  dplyr::mutate(
    # For all comparisons, make sure we aren't comparing NA to NA and getting TRUE
    final_label = dplyr::case_when(
      ########### Check for exact match between SingleR/scANVI
      !is.na(scanvi_label_ontology) & singler_label_ontology == scanvi_label_ontology ~ scanvi_label, 
      ########### Check for family match between SingleR/scANVI
      !is.na(scanvi_family_ontology) & singler_family_ontology == scanvi_family_ontology ~ scanvi_family,
      ########## Now use agreement with consensus to assign a label: first by label, and then by family
      !is.na(consensus_ontology) & singler_label_ontology  == consensus_ontology ~ singler_label,
      !is.na(consensus_ontology) & scanvi_label_ontology   == consensus_ontology ~ scanvi_label,
      !is.na(consensus_family_ontology) & singler_family_ontology == consensus_family_ontology ~ singler_family,
      !is.na(consensus_family_ontology) & scanvi_family_ontology  == consensus_family_ontology ~ scanvi_family,
      # Everything else is Unknown
      .default = "Unknown"
    )
  ) |>
  # Now, we can bring map the final ontologies into the data frame
  dplyr::inner_join(
    ontology_df |> dplyr::rename(final_label = NBAtlas_label, final_id = CL_ontology_id), 
    by = "final_label"
  ) |>
  # add tumor column for visualization
  dplyr::mutate(cell_class = dplyr::case_when(
    final_label == "Neuroendocrine" ~ "tumor", 
    final_label == "Unknown" ~ "unknown", 
    .default = "normal"
  ))
```

@@

```{r fig.width = 10, fig.height = 10}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = final_label) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_manual(values = celltype_pal) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
  umap_theme +
  theme(legend.position = "bottom")
```

@@

```{r fig.width = 8, fig.height = 8}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = cell_class) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_brewer(palette = "Set2") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```

@@

```{r fig.width = 8, fig.height = 8}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = is_xenograft) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_brewer(palette = "Set1") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```

@@

Marker genes that are present in three or more cell types are not included.
In addition, the upper limit for the color scale has been censored to a maximum of 6 to ensure visibility of expression levels across all cell types.

```{r}
input_dotplot_df <- final_annotation_df |>
  dplyr::filter(final_label != "Unknown") |>
  # recode cell types to match what is present in NBAtlas marker genes
  dplyr::mutate(
    label_recoded = dplyr::case_when(
      stringr::str_detect(final_label, "monocyte") ~ "Monocyte", 
      stringr::str_detect(final_label, "cDC") ~ "cDC",
      .default = final_label
  )) |>
  dplyr::select(cell_id, label_recoded)

@@

```{r, fig.width = 32, fig.height = 16, eval = !params$testing}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  celltype_df = input_dotplot_df,
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = celltype_pal, 
  min_cells = 10 # only show cell types with at least 10 cells
) +
  theme(legend.position = "bottom")
```

>>>>

```{r}
final_annotation_df <- annotation_df |> 
  dplyr::mutate(
    # For all comparisons, make sure we aren't comparing NA to NA and getting TRUE
    final_label = dplyr::case_when(
      ########### Check for exact match between SingleR/scANVI
      !is.na(scanvi_label_ontology) & singler_label_ontology == scanvi_label_ontology ~ scanvi_label, 
      ########### Check for family match between SingleR/scANVI
      !is.na(scanvi_family_ontology) & singler_family_ontology == scanvi_family_ontology ~ scanvi_family,
      ########## Now use agreement with consensus to assign a label: first by label, and then by family
      !is.na(consensus_ontology) & singler_label_ontology  == consensus_ontology ~ singler_label,
      !is.na(consensus_ontology) & scanvi_label_ontology   == consensus_ontology ~ scanvi_label,
      !is.na(consensus_family_ontology) & singler_family_ontology == consensus_family_ontology ~ singler_family,
      !is.na(consensus_family_ontology) & scanvi_family_ontology  == consensus_family_ontology ~ scanvi_family,
      # Everything else is Unknown
      .default = "Unknown"
    )
  ) |>
  # Now, we can bring the final ontologies into the data frame
  dplyr::inner_join(
    ontology_df |> dplyr::rename(final_label = NBAtlas_label, final_id = CL_ontology_id), 
    by = "final_label"
  ) |>
  # add tumor column for visualization
  dplyr::mutate(cell_class = dplyr::case_when(
    final_label == "Neuroendocrine" ~ "tumor", 
    final_label == "Unknown" ~ "unknown", 
    .default = "normal"
  ))
```

@@

```{r fig.width = 8, fig.height = 8}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = final_label) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_manual(values = celltype_pal) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
  umap_theme +
  theme(legend.position = "bottom")
```

@@

```{r fig.width = 7, fig.height = 7}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = cell_class) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_brewer(palette = "Set2") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```

@@

```{r fig.width = 7, fig.height = 7}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = is_xenograft) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_brewer(palette = "Set1") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```

@@

Marker genes that are present in three or more cell types are not included.
Only cell types with at least 10 detected cells are shown, and cells without an assigned label are not shown.
In addition, the upper limit for the color scale has been censored to a maximum of 6 to ensure visibility of expression levels across all cell types.

```{r}
input_dotplot_df <- final_annotation_df |>
  dplyr::filter(final_label != "Unknown") |>
  # keep only cells that appear > 10 times
  dplyr::add_count(final_label) |>
  dplyr::filter(n > 10) |>
  # recode cell types to match what is present in NBAtlas marker genes
  dplyr::mutate(
    label_recoded = dplyr::case_when(
      stringr::str_detect(final_label, "monocyte") ~ "Monocyte", 
      stringr::str_detect(final_label, "cDC") ~ "cDC",
      .default = final_label
  )) |>
  dplyr::select(cell_id, label_recoded)

@@

```{r, fig.width = 24, fig.height = 16}
# placing this in eval in chunk settings makes the chunk
# ignore width and height, so we use an if
if (!(params$testing)) {
  generate_dotplot(
    merged_sce = merged_sce,
    markers_df = top_nbatlas_markers_df,
    celltype_df = input_dotplot_df,
    total_cells_df = total_cells_df,
    expressed_genes = expressed_genes,
    bar_order = nbatlas_bar_order, 
    celltype_palette = celltype_pal,
    text_size = 22,
    bar_text_size = 20
  ) +
    theme(legend.position = "bottom")
}

```
